---
title: "CINAR MSE Workshop: Age Structured Model"
author: "Gavin Fay"
date: "January 2022"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(furrr)
library(RcppRoll)
```

```{r}
source("age-functions.R")
source("performance-metrics.R")
```


```{r}
# my_future_options <- future_options()
# my_future_options$globals <- ls()
# my_future_options$packages <- c("tidyverse")
# future::plan(future::multisession)
```


## Age-structured Fisheries Model

$$ N_{1,t+1} = R_{t+1} $$
$$ N_{a+1,t+1} = N_{a,t}e^{-(F_{a,t}+M)} $$

$$ N_{A,t+1} = N_{A-1,t}e^{-(F_{A-1,t}+M)} + N_{A,t}e^{-(F_{A,t}+M)}$$


## Matrix form Age-structured Fisheries Model
\begingroup
\fontsize{7}{7}\selectfont
$$\mathbf{N}_{t+1} = 
\begin{bmatrix}
R_{t+1} \\
0 \\
0 \\
0 \\
\cdots \\
0 \\
\end{bmatrix} +
\begin{bmatrix}
0 & 0 & 0 & \cdots & 0 & 0 \\
e^{-(F_{0,t}+M)} & 0 & 0 & \cdots & 0 & 0 \\
0 & e^{-(F_{1,t}+M)} & 0 & \cdots & 0 & 0 \\
0 & 0 & e^{-(F_{2,t}+M)} & \cdots & 0 & 0 \\
\cdots & \cdots & \cdots & \cdots & \cdots & \cdots \\
0 & 0 & 0 &\cdots & e^{-(F_{A-1,t}+M)} & e^{-(F_{A,t}+M)} \\
\end{bmatrix} \begin{bmatrix}
N_{0,t} \\
N_{1,t} \\
N_{2,t} \\
N_{3,t} \\
\cdots \\
N_{A,t} \\
\end{bmatrix}$$
\endgroup




```{r}
# # ## read in biol data & create input object
#path <- "materials/exercises/day-02/"
input <- list()
input$nages <- scan("floundah_biology.txt",n=1,skip=9)
nages <- input$nages
input$maturity <- scan("floundah_biology.txt",n=nages,skip=11)
input$selex <- scan("floundah_biology.txt",n=nages,skip=13)
input$weight <- scan("floundah_biology.txt",n=nages,skip=15)
#
input$M <- 0.4
input$Fmult <- 1 #proportion of FSPRtarg during historical period
input$SPRtarg <- 0.4
input$Rbar <- 1000
input$sigmaR <- 0.6
input$cpueCV <- 0.3
#xx <- evaluate(input, seed=24601)
```


```{r}
om_settings <- tibble(om=1,
                      input=list(input)) %>% 
  I()
om_settings$input
om_settings
```
set up the random number seeds for each iteration
```{r}
nsim <- 1000
set.seed(24601)
seeds <- ceiling(99999*runif(nsim))

sim_seeds <- tibble(isim = 1:nsim,
                    seed = seeds)
```


set up the table for running the mse over all the operating models (just 1 in this example)
```{r}
mse_sim_setup <- om_settings %>% 
  mutate(sim_seeds= list(sim_seeds)) %>% 
  unnest(cols = sim_seeds) %>% 
  rowid_to_column() %>% 
  select(rowid,everything()) %>% 
  ungroup() %>% 
  arrange(isim) %>%   # organizes so first realization of each is done first
  I()
mse_sim_setup
```

run the mse for all the iterations
```{r}
### run the MSE over each row of the mse_sims todo
#add a safe mode (returns error safely rather than crashing)
safe_evaluate <- purrr::safely(evaluate, otherwise = NA_real_)
#do the MSE for all simulations and scenarios
#profvis::profvis(
  #system.time(
mse_output <- mse_sim_setup %>% 
#   mutate(mse = furrr::future_pmap(list(seed = seed, input = input),
#                           safe_evaluate, .options = my_future_options)) %>% 
# this is the regular purrr code for iterating over the simulations
mutate(mse = purrr::pmap(list(seed = seed, input = input), evaluate)) %>% 
  rowwise() %>%
  mutate(om_SSB = list(mse$om_SSB)) %>%
  mutate(om_F = list(mse$om_F)) %>%
  mutate(tac = list(mse$tac)) %>%
  mutate(ftarg = mse$ftarg) %>%
  I()
mse_output
```


```{r}
mse_results <- mse_output %>% 
  rowwise() %>% 
  mutate(bmsy = input$Rbar*SBPR(ftarg,input$M,input$weight,input$selex,input$maturity)) %>% 
  mutate(msy = input$Rbar*YPR(ftarg,input$M,input$weight,input$selex)) %>% 
  ungroup() %>% 
  mutate(ssb_metrics = pmap(list(om_SSB, bmsy), get_ssb_metrics, nprojyrs = 50),
         catch_metrics = pmap(list(tac, msy), get_catch_metrics, nprojyrs = 50),
         f_metrics = pmap(list(om_F, ftarg), get_F_metrics, nprojyrs = 50)) %>% 
  select(rowid, om, isim, ssb_metrics, catch_metrics, f_metrics) %>% 
  I()
mse_results
```

`mse_results` is a tibble containing the results, and a list of SSB, catch, and F performance metrics

pull out the  metrics

```{r}
ssb_results <- mse_results %>% 
  select(rowid, om, isim, ssb_metrics) %>% 
  mutate(ssb_metrics = map(ssb_metrics, enframe)) %>% 
  unnest(cols = c(ssb_metrics)) %>% 
  mutate(value = map_dbl(value, I)) %>% 
  rename(metric = name) %>% 
  I()
ssb_results
unique(ssb_results$metric)

###pull out the f metrics
f_results <- mse_results %>% 
  select(rowid, om, isim, f_metrics) %>% 
  mutate(f_metrics = map(f_metrics, enframe)) %>% 
  unnest(cols = c(f_metrics)) %>% 
  mutate(value = map_dbl(value, I)) %>% 
  rename(metric = name) %>% 
  I()
f_results
unique(f_results$metric)

###pull out the catch metrics
catch_results <- mse_results %>% 
  select(rowid, om, isim, catch_metrics) %>% 
  mutate(catch_metrics = map(catch_metrics, enframe)) %>% 
  unnest(cols = c(catch_metrics)) %>% 
  mutate(value = map_dbl(value, I)) %>% 
  rename(metric = name) %>% 
  I()
catch_results

full_metrics <- bind_rows(ssb_results, catch_results, f_results)
```



summarize across simulations by OM scenario
25%, 50%, 75% quantiles
```{r}
quibble <- function(x, q = c(0.25, 0.5, 0.75)) {
  tibble(x = quantile(x, q, na.rm = TRUE), q = q)
}

summary <- full_metrics %>% 
  group_by(metric, om) %>% 
  summarise(y = list(quibble(value, c(0.25, 0.5, 0.75)))) %>% 
  tidyr::unnest(y) %>% 
  I()
summary
```

